\\Large Rising temperatures alter how and where mammals move

### 2025-09-29 writing/manuscript.Rmd
reviewed intro and abstract

<!--
six-sentence abstract:

1. introduce the topic
2. state the problem
3. summarize why the question hasn't been answered/what the consequences are
4. explain how you answered the question
5. explain the central findings
6. explain the importance of your findings and their consequences
-->

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature significantly affects of how and where these mammals move, and we predict that climate change will have significant impacts on mammals' movement behaviour throughout the 21^st^ century. Notably, we found that species did not respond uniformly to temperature. This suggests that community-level responses will be complex, as future climate change drives some species to move more and others less. Median habitat selection strength was predicted to decrease for most species, but some species are expected to show increased selection strength for higher altitudes. Consequently, our results suggest that individuals from all six species may shift their range in the coming decades as they begin to select against portions of their current ranges. The strong but varied responses to temperature suggest that future climate change will impact not only individuals' movement behaviour but also entire communities, as changes in movement behaviour will affect encounter rates, including predator prey-dynamics and human-wildlife interactions. As climate change exposes mammals to novel environmental conditions, proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

@@

\noindent \textbf{Figures}: 8

@@

\noindent \textbf{References}: 165 (updated on 2025-09-25) <!-- counted by using movement ecology csl -->

@@

<!-- using "behaviOR" in key words to increase audience -->

\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behavior, habitat selection

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. How and where animals move thus determines encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates [@dyer_travel_2023] and even cause physiological damage and death [@hetem_responses_2014; @ratnayake_forecasting_2019; @schmidt_interplay_2020]. Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016]. Changes in climate throughout the 21^st^ century and the simultaneous increase in frequency and intensity of extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011] are expected to impact not only animals' movement behavior but also their physiology [@powers_hovering_2017; @schwerdt_ecological_2024], life history [@newediuk_climate_2024], and phenology [@carbeck_adaptation_2022; @johnson_human_2018].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size. Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].<!-- Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].-->

<!-- we do not know how temperature will change -->

In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023]. As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018]. Preparing for and responding to future changes will require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021]. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021]. However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of changes in movement behaviour with climate change in the current century [@ipcc_climate_2023].

@@

***CLEAN UP PARAGRAPH BELOW***

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023]. Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024]. We discuss the consequences of our findings at the individual, population, and community levels, and we also provide directions for future research and solutions for overcoming some common data limitations.

@@

>>>>

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023. We show that temperature significantly affected of how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century, although species did not respond uniformly to temperature. Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less. Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology. We also predict that rising temperatures will reduce median habitat selection strength most of the six species, but some are expected to show increased selection strength for higher altitudes. As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions. Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities. Proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

\noindent \textbf{Running Title}: Temperature alters mammal movement behavior

@@

\noindent \textbf{Figures}: 9

@@

\noindent \textbf{References}: 163 (updated on 2025-09-29) <!-- counted using movement ecology csl -->

@@

\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behaviour, habitat selection, projections

\noindent \textbf{Contact Information:} *Corresponding author’s telephone, and email details*

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024]. Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016]. However, climate change during the last century [@ipcc_climate_2023] have caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011].

@@

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024]. Mammals' homeothermy can cause them to have difficulty dissipating heat, particularly when ambient temperatures are near or above body temperature. Large-bodied mammals are most likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023]. Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024]. However, behavioral responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018]. For example, warming temperatures have allowed white-tailed deer (*Odocoileus virginianus*) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024]. Preparing for and responding to future changes will require a detailed understanding of not only how temperature affects how and where mammals move but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].

<!-- responding to future changes -->

As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020]. It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @botero_evolutionary_2015; @williams_range_2018]. Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021], but there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of future changes in movement behaviour.

@@

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023]. Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024]. We discuss the consequences of our findings at the individual, population, and community levels, and we also provide directions for future research and solutions for overcoming some common data limitations.

@@

Points to add:

* thermal landscape?
* comon trends:
  * boreal species do not see much of a change in distance traveled, but this is in part due to opposite trends in the different seasons.
  * habitat selection for boreal species flattens with temperatures above 0 but strengthens near zero. Therefore, warmer summers may result in lower encounter rates as ranges widen, but warmer winters may result in increased predation risk, which may impact reproductive success (since cows are pregrant in winter)
  * SM caribou and mountain goats are projected to move more and see widespread decreases in their current habitats
